Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-04-13 WholePlotTrajectories.Rmd
Files

>The distance from the initial condition was significantly time-dependent (P<0.01).
>The coordinates of the functional traits were measured in the two-dimensional ordination space that mapped plots evolution along time (graphs not showed) which revealed that the trajectories of disturbed plots along time headed toward acquisitive functional strategies (from high SWG to high SLA and chlorophyll content).

@@

```{r Fig4, echo=FALSE, out.width='100%', fig.asp=2/5,fig.cap="Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/Smass_ClassProp")

>>>>

The coordinates of the functional traits were measured in the two-dimensional ordination space that mapped plots evolution along time (graphs not showed) revealed a trajectory of disturbed plots toward acquisitive functional strategies (from high WD to high SLA and chlorophyll content).

@@

```{r Fig4, echo=FALSE, out.width='100%', fig.asp=2/5,fig.cap="Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure*"}
load("DB/Smass_ClassProp")

### 2018-04-13 WholePlotTrajectories.tex
Files

\begin{quote}
The distance from the initial condition was significantly time-dependent
(P\textless{}0.01). The coordinates of the functional traits were
measured in the two-dimensional ordination space that mapped plots
evolution along time (graphs not showed) which revealed that the
trajectories of disturbed plots along time headed toward acquisitive
functional strategies (from high SWG to high SLA and chlorophyll
content).
\end{quote}

@@

\caption{Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.}\label{fig:Fig4}
\end{figure*}

@@

{\centering \includegraphics{WholePlotTrajectories_files/figure-latex/Fig6-1}

@@

In line with the ``vegetation quantity effect'' suggesting that
communities functional characteristics depends on dominant species
rather that subtle changes in biodiversity \citep{Grime1998}. This would
result in a faster recovery of ecosystem processes rates than
composition and diversity as demonstrated for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. This highlighted the functional
overlap between species in tropical forests, typical of their huge
biodiversity, and intricately questioned the different facets of
communities resilience. Ecosystem processes grasped through the set of
functional traits retained here however do not assess communities
resilience and future dynamics.

In line with the assumtions of the IDH the observed diversity
trajectories confirmed the increase of communities evenness and
functional richness after disturbance.
\textless{}\textless{}\textless{}\textless{}\textless{}\textless{}\textless{}
HEAD The consistency of the IDH has been debated for the higly diverse
tropical forests, some observing the importance of environmental filters
to shape communities assembly and others arguing for a genuine
stochastic recovery. Consistently to the trajectories drawn here, a
previous study of the same experiment revealed the directional changes
in communities composition reflecting environmental filtering after
disturbance. Disturbance regimes brought a unimodal response depending
on disturbance intensity as suggested by the outcome of 10 years and 20
years post-disturbance
analysis\citep{Hubbell1999, Molino2001, Baraloto2012a}. As observed
after logging in a Boranean rainforest \citep{Cannon1998} or after
cyclones in the Solomon islands \citep{Burslem2000}, communities
diversity showed a unimodal response to disturbance and a maintenance of
diversity levels, except for one of the intensively disturbed community
which evenness was still increasing \ref{fig:Fig5}. Despite the recovery
of communities diversity, though, taxonomic composition was clearly
altered \ref{fig:Fig3} implying a long-term, if not permenant,
composition change after disturbance as observed in East African
tropical forests \citep{Bonnell2011}.

Regarding the functional trajectories and dominant trait values of
disturbed communities (Figure \ref{fig:Fig3}), disturbance appeared to
favor acquisitive strategies, \emph{i.e} species with high SLA, low wood
density, small seeds, etc \citep{Westoby1998, Wright2004, Reich2014}.
The comparison of observed trajectories to a null model with random
traits set shuffling showed an oriented functional shift after
disturbance which suggest additional biotic and abiotic filters for the
10 to 15 years after disturbance. This means a shift from exclusive
competition favoring competitive species before disturbance to
environmental filtering selecting acquisitive and disturbance resistant
species. In particular, the evolution seed mass classes proportions
(Figure \ref{fig:Fig4}) revealed a clear increase in small seeded,
therefore large dispersive, species
\citep{TerSteege2001, Flores2006, Haddad2008}.

The diversity and composition functional trajectories showed a
decoupling between functional and taxonomic after the most intense
disturbance (treatments T2 and T3). Plots functional characteristic
returned to their initial values, similar to those of control plots,
faster than their taxonomic counterparts (Figures \ref{fig:Fig2} to
\ref{fig:Fig4}). As functional traits are the most direct link between
biodiversity and ecosystem functioning \citep{Diaz2005} the recovery of
communities functional characteristics would go along the recovery of
ecosystems processes although taxonomic composition remain altered
\citep{Guariguata2001}. Such decoupling suggested the consistency of
functional overlap and maybe functional versatility, implying a variety
of functioning from the same characteristic, in the pre-disturbance
condition. Functional overlap has been observed in highly diverse
ecosystems \citep{Bellwood2006} and suggested as one mechanisms for high
diversity maintenance. Knowing the decoupling between taxonomic and
functional trajectories, this functional overlap may not be recovering
after disturbance and therefore radically modify communities resilience
and resistance \citep{Trenbath1999, Elmqvist2003, Diaz2005}. Given the
impaired functional overlap, the persitant compositional changes
regarding disturbance resistant species \citep{Haddad2008} and probably
lianas or epiphytes \citep{Martin2013}, and the modification of
environmental conditions like soils nutrient cycling and compaction
\citep{Olander2005} the entirety of communities recovery after intense
disturbance is deeply questioned \citep{Chazdon2003a}. New conditions
would not only be longer lasting but self-maintained as tied to
disturbance regime \citep{Burslem2000}. Although communities recovered
the core of their initial functioning, composition seemed permanently
altered after intense disturbance. Specificaly, this would impair
species contingent to undisturbed forests, threatening their
maintenance, and run the risk to loose cornerstone species and trigger
unexpected ecological consequences
\citep{Jones1994, Diaz2005, Gardner2007}.

@@

\begin{quote}
Important differences among plots of same treatments: possible impact of
initial diversity composition, local abiotic parameters (topography,
soils, surrounding stands). Impacts depend on the disturbance intensity,
but further parameters may be involved.
\end{quote}

>>>>

The coordinates of the functional traits were measured in the
two-dimensional ordination space that mapped plots evolution along time
(graphs not showed) revealed a trajectory of disturbed plots toward
acquisitive functional strategies (from high WD to high SLA and
chlorophyll content).

@@

\caption{Trajectories of seed mass classes proportions over 30 years after disturbance, corresponding to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. No gap filling process was applied in this case. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:Fig4}
\end{figure*}

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/Fig6-1}

@@

\subsection{Validation of the IDH}\label{validation-of-the-idh}

The enhanced communities evenness and functional richness after
disturbance confirmed the IDH for all treatment intensity. The
divergence from the initial state however was very variable among plots,
specifically after intense disturbance, which meant a strong dependency
of plots recovery to initial composition or to abiotic conditions. All
plots, though, followed the same unimodal trajectory which confirmed
both the resilience of forests taxonomic and functional composition and
the convergence of communities towards initial states
\citep{Hubbell1999, Molino2001, Baraloto2012a}. Functional traits are
the most direct link between biodiversity and ecosystem functioning
\citep{Diaz2005} the recovery of communities functional characteristics
would go along the recovery of ecosystems processes although taxonomic
composition remain altered \citep{Guariguata2001}. Disturbance proved to
favor species with acquisitive strategies, displaying efficient resource
acquisition (high SLA) but short-lived tissues (low wood density, leaf
toughness and thickness) and with a dispresive strategy (low seed mass)
\citep{Westoby1998, Wright2004, Reich2014}. \#The comparison of observed
trajectories to a null model with random traits set shuffling showed an
oriented functional shift after disturbance which suggest additional
biotic and abiotic filters for the 10 to 15 years after disturbance.
\#This means a shift from exclusive competition favoring competitive
species before disturbance to environmental filtering selecting
acquisitive and disturbance resistant species. In particular, the
evolution seed mass classes proportions revealed a clear increase in
small seeded, therefore large dispersive, species
\citep{TerSteege2001, Flores2006, Haddad2008}.

\subsection{Functional redundancy of disturbed
ecosystems}\label{functional-redundancy-of-disturbed-ecosystems}

After disturbance, communities functional characteristics followed a
similar but delayed trajectory as this of taxonomic characteristics
which difference to initial state remained higher after 30 years. This
confirms conclusions drawn for grasslands
\citep{Tilman1997, Mouillot2011} and more recently for tropical forests
\citep{Lohbeck2015, Guariguata2001}. Assuming the ``vegetation quantity
effect'', suggesting that communities functional characteristics are
defined by dominant species rather that subtle changes in biodiversity
\citep{Grime1998}, this delay between functional and taxonomic
diversities meant the rapid recovery of dominant species while this of
rare ones would be much longer. Besides, it meant that species missing
to plots taxonomic recovery would be functionally redundant with those
already in place. The functional overlap between species, typical of the
huge biodiversity of tropical forests \citep{Bellwood2006}, is then not
trivially recovered after disturbance and as a determinant of forests
resilience this questions forests resilience to multiple disturbance
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Besides such impact on functional redundancy along with persitant
compositional changes regarding disturbance resistant species
\citep{Haddad2008} and probably lianas or epiphytes \citep{Martin2013},
and the modification of environmental conditions like soils nutrient
cycling and compaction \citep{Olander2005} forest resilience remains
under question \citep{Chazdon2003a}. New conditions would not only be
longer lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.

@@
